\documentclass{article}
\setlength\parindent{0pt}
\usepackage{mathtools}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}

\title{Build Heap: Insertion vs Bottom-Up\\(Runtime Proof)}
\author{Casey Li}

\begin{document}
\maketitle

\part*{Insertion (topdown) - insert, bubble, repeat}
Assuming max-heap (biggest items at top): \\

Best case: $\theta$(n) \\
Worst case: $\theta$(nlogn) \\
Average case: $\theta$(n)\\

\section*{Best case proof}
Items in descending order. Biggest are inserted first.\\
Insertion occurs n times.\\
Each one bubbles exactly 0 times.

\section*{Worst case proof}
Items come in ascending order.\\
Insert occurs n times.\\
For each insertion, the item must bubble to the top of the tree.\\
At any given insertion of item i, the height of the tree is $\floor{logn}$.\\
Total nlogn times.

\section*{Average proof}
\subsection*{Upper bound, O(n)}
1 + 2 + 4 + 8 = 15.

For a complete binary tree of 15 items (almost 15): \\

8/15 items bubble 0 times ($\approx$ 8/16 $\approx$ 1/2, strictly $<$ 1) \\
4/15 items bubble 1 times ($\approx$ 4/16 $\approx$ $1/2^2$, $<$ 1/2) \\
2/15 items bubble 2 times ($\approx$ 2/16 $\approx$ $1/2^3$ $<$ 1/4)\\
1/15 items bubble 3 times ($\approx$ 1/16 $\approx$ $1/2^4$ $<$ 1/8)\\

P(bubble i times): $\dfrac{i}{2^{i+1}}$, which is strictly less than  $\dfrac{i}{2^{i}}$\\

Expected bubbling of any given item is less than (upper bound):

$$
\sum_{i=0}^{\floor{logn}} \dfrac{i}{2^{i}}
$$

This sum converges to 2 as i approaches infinity, so this particular sum is strictly less than 2. In other words, it is constant, $\theta$(n). \\

We then have upper bound on the average = 2n = $O$(n).

\subsection*{Lower bound, $\Omega$(n)}
So now we just need to pick a case to prove that in the best case, the average running time is $\Omega$(n).\\

Easy. We already did this. Best case, they are inserted in descending order, each one bubbles 0 times, so we still have $\Omega$(n).\\

Having proved both the upper and lower bounds, we can conclude that the average case runs in $\theta(n)$.
\newpage
\part*{Bottom-up: fill all, then fix subheaps}

In any given heap, \\

$1/2$ of the items are leaves, no fixing \\
$1/2^2$ of the items are heaps of height 1 (max bubble 1 level) \\
$1/2^3$ of the items are heaps of height 2 \\
$1/2^4$ of items are heaps of height 3 \\

In general, $1/2^{h+1}$ items will bubble up at most h times. \\

Worst case \# of bubbles in a filled binary tree is:

$$n \sum_{i=0}^{\floor{logn}} \dfrac{h}{2^{h+1}} = 
\dfrac{n}{2} \sum_{i=0}^{\floor{logn}} \dfrac{h}{2^{h}}$$ \\

The summation converges to 2 as n goes to infinity, so this sum is strictly less than n. So worst case, this runs in $\theta$(n) time.

\end{document}